---
title: "Data 607 - Week 6 - Project 2"
subtitle: "Cosmetics Ingredients"
author: "Glen Dale Davis"
date: "2023-02-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages:

Below, the packages required for data analysis and visualization are loaded.

```{r packages, warning = FALSE, message = FALSE}
library(tidyverse)
library(magrittr)
library(DT)

```

## Cosmetics Ingredients Data Collection:

We load data related to cosmetics Products from a number of Brands and those Products' Ingredients. We remove a number of columns not needed for analysis.

```{r cosmetics1}
my_url1 <- "https://raw.githubusercontent.com/geedoubledee/data607_project2/main/cosmetics.csv"
cosmetics <- read.csv(my_url1)
cosmetics_new <- subset(cosmetics,
                    select = -c(Rank, Combination, Dry, Normal,
                                Oily, Sensitive))

datatable(head(cosmetics_new), options = list(pageLength = 3))

```

## Cosmetics Ingredients Data Clean Up:

Most of the Ingredients data for each Product is stored as a character string, delimited by a comma, so we separate those strings into their individual components and widen this one column into multiple columns. We name the new columns starting with "Ingredients_", followed by that Ingredient's integer position in the original comma-delimited string we separated (1, 2, etc.). We fill in NAs as needed if a Product's number of Ingredients doesn't match the number of "Ingredients_" columns we've created, as all Products don't have the same number of Ingredients.

```{r cosmetics2}
cosmetics_new %<>%
    separate_wider_delim(Ingredients, delim = ", ", names_sep = "_",
                         too_few = "align_start")

datatable(head(cosmetics_new), options = list(scrollX = TRUE))

```

Then we pivot this wide data into a longer format where an Ingredient's integer position in the original comma-delimited string becomes a variable called Listed_Order, and the matching Ingredient goes in the Ingredient column.

```{r cosmetics3}
cosmetics_new %<>%
    pivot_longer(cols = starts_with("Ingredients_"),
                 names_to = "Listed_Order",
                 values_to = "Ingredient") %>%
    filter(!is.na(Ingredient))

cosmetics_new$Listed_Order <- as.integer(
    str_replace_all(cosmetics_new$Listed_Order, "Ingredients_", ""))

datatable(head(cosmetics_new))

```

## Cosmetics Ingredients Data Analysis:

We now count the number of Products each Ingredient is used in, and we sort the results in descending order. We display the top 25 most used Ingredients across all Brands.

```{r cosmetics4}
cosmetics_new_analysis <- cosmetics_new
cosmetics_new_analysis %<>%
    group_by(Ingredient) %>%
    summarize(Products = n()) %>%
    arrange(desc(Products))

datatable(head(cosmetics_new_analysis, 25))

```
